Higher-Order unification for free!

### 2024-04-17 paper/main.tex
update scope-check

\subsubsection{Scope check}\label{sec:sc}
The predicate xxxx \todo{add} takes
a link \linketaM{\Gamma}{X_{x_1\ldots x_n}}{t} and replaces it by
\linketaM{{x_1\ldots x_n}}{X_{x_1\ldots x_n}}{t'} whenever $\{x_1\ldots x_n\} \subset \Gamma$.
The term $t' = \lambda x.Y_{{x_1\ldots x_n},x}$ after
executing $\lambda x.Y_{{x_1\ldots x_n}~x} \Ue t$, that is $t'$ cannot contain
variables in $\Gamma / \{x_1\ldots x_n\}$.
This is key for fidelity since suspended unif problem
must stay solvable.

@@

\newcommand{\progressetaright}{\emph{$\eta$-progress-\rhs}\xspace}
\begin{definition}[\progressetaright]\label{def:progressetaright} A link
\linketaM{\Gamma}{X}{t} is removed from \linkStore when either 1)
$\testmaybeeta~t$ does not hold (anymore) or 2) by $\eta$-contracting $t$ to
$t'$ where $t'$ does not start with the \elpiIn{lam} constructor. In the
first case, $X$ is unified with $t$, and in the second one, $X$ is unified with
$t'$ (under the context $\Gamma$).
\end{definition}
\todo{comment this rule}

@@

If we look back the example give at the beginning of this section, we can
deduplicate $\mapping{X}{E}{1}, \mapping{X}{C}{2}$ by removing the first mapping
and adding the auxiliary \linketa: \linketaM{x}{E_{x}}{\lambda y.C_{x y}}.
After deduplication the compiler output is as follows:
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{Y,F,0},
    {X,C,2}}}
  {{{\eta,x,E_{x},\lambda y.C_{x y}},
    {\eta,,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {\eta,,A,\lambda x.B_{x}},
    {\eta,x,B_{x},\lambda y.C_{y x}}}}

\noindent
The resolution of $\hoUnifPb_1$ assigns $\lambda x.\lambda y.x$ to $A$
that in turn triggers $\linkStore_3$ and then $\linkStore_4$ by \progressetaleft,
one per $\lambda$ in the solution of $A$.
The unification variable $C_{yx}$ is therefore unified with $x$
(the second variable of its scope).
As a result $\linkStore_1$ becomes \linketaM{x}{E_x}{\lambda y.y},
and the \rhs is no more in \maybeeta so \progressetaright fires and
$E_x$ is unified with $\lambda y.y$; then $\linkStore_2$ makes progress thanks to the same 
rule giving
$\sigma = \{~ A \mapsto \lambda x.\lambda y.x ~;~ B_x \mapsto \lambda y.x ~;~ C_{yx} \mapsto x ~;~ D \mapsto f\appsep (\lambda y.y) ~;~ E_x \mapsto \lambda y.y ~\}$.\\
The remaining step $\hoUnifPb_2$ identifies $D$ with $F$,
hence the resulting \Fo{} substitution is
$\rho = \{X \mapsto \lambda x.\lambda y.y ~;~ Y \mapsto f\appsep (\lambda y.y)\}$.

@@

\begin{definition}[\progBetaLL]
  Let $\sigma$ be a substitution such that $\sigma t_1$ is a name $s$
  not occurring$s_1\dots s_n$. If $m = 1$, then $l$ is removed and \lhs is
  unified with $X_{s_1\ldots s_n~s}$.
  If $m > 1$, then $l$ is replaced by the link
  $\linkbetaM{\Gamma}{T}{Y_{s_1\dots s_n~s}\appsep t_2\dots t_m}$
  (where $Y$ is a fresh variable of arity $n+1$) and link
  \linketaM{\Gamma}{X_{s_1\dots s_n}}{\lambda x.Y_{s_1\dots s_n~x}}
  is added to \linkStore.

  % A link \linkbetaM{\Gamma}{T}{\rhsBeta} is removed from \linkStore, if given
  % the substitution $\sigma$, $\sigma t_1$ is a name, say $t$, such that, $t
  % \notin s_1\dots s_n$. In this case, let $Y$ a fresh variable, then
  % \linketaM{\Gamma}{X_{s_1\dots s_n}}{\lambda x.Y_{s_1\dots s_n,x}} is added to
  % \linkStore, and 1) if $m = 1$ then $Y_{s_1\dots s_n, t}$ is unified with $T$,
  % 2) otherwise, the refined \linkbeta, \linkbetaM{\Gamma}{T}{M_{s_1\dots
  % s_n,t}\appsep t_2\dots t_m}, is added to \linkStore.
  \label{def:progBetaLL}
  % % \noteE[inline]{non è chiaro. la sostituzione immagino sia la corrente, non se ne esiste una. Poi non è chiaro se otherwise si riferische a in this case. Usa un enumerate}
  % %\noteE[inline]{Is it clearer?}
  % \noteE[inline]{l1 or t1? Forse è più chiaro dire che il ilnk beta viene raffinato con link beta cone meno argomenti, Se poi il numero di argomenti extra è 0 il tutto è in llambda e quindi viene buttato via. In questo modo la terminazione è ancora più chiara perchè si vede già che prima decresce la lista di argomenti e poi il numero di beta}
  % \noteD{L'ho riformulato}
\end{definition}

@@

%\input{code/progress_heuristic}
%\noindent
%Note that \elpiIn{UV} is in \llambda as well as \elpiIn{T}. 
\vspace{-1em}
\section{Actual implementation in Elpi}\label{sec:implementation}

>>>>

\subsubsection{Scope check}\label{sec:sc}
The predicate \elpiIn{scope-check} walks through all links. 
Each link of the form \linketaM{\Gamma}{X_{x_1\ldots x_n}}{t} is replaced by
\linketaM{{x_1\ldots x_n}}{X_{x_1\ldots x_n}}{t'} whenever $\{x_1\ldots x_n\} \subset \Gamma$.
The term $t' = \lambda x.Y_{{x_1\ldots x_n},x}$ is obtained after
executing $\lambda x.Y_{{x_1\ldots x_n}~x} \Ue t$. This unification
ensures that $t'$ cannot contain variables in $\Gamma / \{x_1\ldots x_n\}$,
and therefore grants fidelity.
In fact, a link represents a suspended unification problem between a unification variable $X$
and a ``problematic'' term $t$. This unification is possible iff 
the free variables in $t$ are also in the scope of $X$.

@@

\begin{comment}
\newcommand{\progressetaright}{\emph{$\eta$-progress-\rhs}\xspace}
\begin{definition}[\progressetaright]\label{def:progressetaright} A link
\linketaM{\Gamma}{X}{t} is removed from \linkStore when either 1)
$\testmaybeeta~t$ does not hold (anymore) or 2) by $\eta$-contracting $t$ to
$t'$ where $t'$ does not start with the \elpiIn{lam} constructor. In the
first case, $X$ is unified with $t$, and in the second one, $X$ is unified with
$t'$ (under the context $\Gamma$).
\end{definition}
\todo{comment this rule}
\end{comment}

@@

If we look back the example give at the beginning of this section, we can
deduplicate $\mapping{X}{E}{1}, \mapping{X}{C}{2}$ by removing the first mapping
and adding the auxiliary \linketa: \linketaM{x}{E_{x}}{\lambda y.C_{x y}}.

After deduplication the compiler output is as follows:\todo{add simple example}
\begin{comment}
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{Y,F,0},
    {X,C,2}}}
  {{{\eta,x,E_{x},\lambda y.C_{x y}},
    {\eta,,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {\eta,,A,\lambda x.B_{x}},
    {\eta,x,B_{x},\lambda y.C_{y x}}}}

\noindent
The resolution of $\hoUnifPb_1$ assigns $\lambda x.\lambda y.x$ to $A$
that in turn triggers $\linkStore_3$ and then $\linkStore_4$ by \progressetaleft,
one per $\lambda$ in the solution of $A$.
The unification variable $C_{yx}$ is therefore unified with $x$
(the second variable of its scope).
As a result $\linkStore_1$ becomes \linketaM{x}{E_x}{\lambda y.y},
and the \rhs is no more in \maybeeta so \progressetaright fires and
$E_x$ is unified with $\lambda y.y$; then $\linkStore_2$ makes progress thanks to the same 
rule giving
$\sigma = \{~ A \mapsto \lambda x.\lambda y.x ~;~ B_x \mapsto \lambda y.x ~;~ C_{yx} \mapsto x ~;~ D \mapsto f\appsep (\lambda y.y) ~;~ E_x \mapsto \lambda y.y ~\}$.\\
The remaining step $\hoUnifPb_2$ identifies $D$ with $F$,
hence the resulting \Fo{} substitution is
$\rho = \{X \mapsto \lambda x.\lambda y.y ~;~ Y \mapsto f\appsep (\lambda y.y)\}$.
\end{comment}

@@

\begin{definition}[\progBetaLL]
  Let $\sigma$ be a substitution such that $\sigma t_1$ is a name $s$
  not occurring$s_1\dots s_n$. If $m = 1$, then $l$ is removed and \lhs is
  unified with $X_{s_1\ldots s_n~s}$.
  If $m > 1$, then $l$ is replaced by the link
  $\linkbetaM{\Gamma}{T}{Y_{s_1\dots s_n~s}\appsep t_2\dots t_m}$
  (where $Y$ is a fresh variable of arity $n+1$) and link
  \linketaM{\Gamma}{X_{s_1\dots s_n}}{\lambda x.Y_{s_1\dots s_n~x}}
  is added to \linkStore.   
  %
  % A link \linkbetaM{\Gamma}{T}{\rhsBeta} is removed from \linkStore, if given
  % the substitution $\sigma$, $\sigma t_1$ is a name, say $t$, such that, $t
  % \notin s_1\dots s_n$. In this case, let $Y$ a fresh variable, then
  % \linketaM{\Gamma}{X_{s_1\dots s_n}}{\lambda x.Y_{s_1\dots s_n,x}} is added to
  % \linkStore, and 1) if $m = 1$ then $Y_{s_1\dots s_n, t}$ is unified with $T$,
  % 2) otherwise, the refined \linkbeta, \linkbetaM{\Gamma}{T}{M_{s_1\dots
  % s_n,t}\appsep t_2\dots t_m}, is added to \linkStore.
  \label{def:progBetaLL}
  % % \noteE[inline]{non è chiaro. la sostituzione immagino sia la corrente, non se ne esiste una. Poi non è chiaro se otherwise si riferische a in this case. Usa un enumerate}
  % %\noteE[inline]{Is it clearer?}
  % \noteE[inline]{l1 or t1? Forse è più chiaro dire che il ilnk beta viene raffinato con link beta cone meno argomenti, Se poi il numero di argomenti extra è 0 il tutto è in llambda e quindi viene buttato via. In questo modo la terminazione è ancora più chiara perchè si vede già che prima decresce la lista di argomenti e poi il numero di beta}
  % \noteD{L'ho riformulato}
\end{definition}

@@

%\input{code/progress_heuristic}
%\noindent
%Note that \elpiIn{UV} is in \llambda as well as \elpiIn{T}. 
% \vspace{-1em}
\section{Actual implementation in Elpi}\label{sec:implementation}
